---
title: "Machine Learning Course Project. Human Activity Recognition"
author: "Andrey Vlasenko"
date: "10.04.2017"
output: html_document
---

# Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 
In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here  <http://groupware.les.inf.puc-rio.br/har>.

The goal of the project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

# Summary
In this work for classification task we take into account 5 different models:
- random forest (rf), 
- decision tree (rpart), 
- stochastic gradient boosting (gbm), 
- linear discriminant analysis (lda), 
- support vector machine (svm).

There was 2 input sets: 
- training (pml-training.csv) which was split to train `trnSet_trn` (biulding the models) and test sets `trnSet_tst` (cross-validation).
- testing (pml-testing.csv) which was used in final quiz. 

To compare models we estimate out of sample error (accuracy). The best result demonstrate RF (0.99), GBM (0.96) and SVM (0.93) models.

```{r,results='hide',message=FALSE,warning=FALSE}
library(caret); library(randomForest); 
library(rpart); library(rpart.plot); library(plyr)
library(gbm); library(e1071); library(MASS)
```

# Data preprocessing
Loading and clearing the data.
```{r}
training<-read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",
                   na.strings=c("", "NA", "NULL", "#DIV/0!")) # there are NA values in the data
testing<-read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",
                   na.strings=c("", "NA", "NULL", "#DIV/0!")) # there are NA values in the data
c(dim(training),dim(testing)) # dimensions of training and testing sets without NA vars
table(training$classe) # there are 5 outcomes.
NAVals <- apply(is.na(training),2,sum)/dim(training)[1] # share of NA values in data
NAVals <- NAVals[NAVals>0.95]; NAnms<-names(NAVals); # NAnms - names of useless variables
trnSet <- training[,!(names(training) %in% NAnms)] # removing NA variables from training
tstSet <- testing[,!(names(training) %in% NAnms)] # removing NA variables from testing
trnSet <- trnSet[,-c(1:7)] # we have to remove first 7 variables linked with time 
tstSet <- tstSet[,-c(1:7)] # and number of the observations to avoid overfitting of the model
c(dim(trnSet),dim(tstSet)) # dimensions of training and testing sets without NA vars
str(trnSet)
ValsOfVars<-apply(trnSet, 2, function(x)length(unique(x))) # Number of values for each variable
length(ValsOfVars[ValsOfVars<2]) # there are no meaningless variables in data (with only 1 value)
```
There are `r dim(trnSet)[2]` variables (`r dim(trnSet)[2]-1` predictors) in training (`r dim(trnSet)[1]` rows) and testing (`r dim(tstSet)[1]` questions of the final test) data sets.

# Modeling

Slicing training data set to training and testing sets to estimate the quality (accuracy) of the models by cross-validation.

```{r}
set.seed(1435)
inTrain <- createDataPartition(y=trnSet$classe, p = 0.60, list=FALSE)
trnSet_trn <- trnSet[inTrain,] # training set (to biuld the models)
trnSet_tst <- trnSet[-inTrain,] # testing set for cross-validation (to estimate error)
```

Let's fit the models (using `trnSet_trn`). We'll take into account the following models: 
* random forest, 
* decision tree, 
* stochastic gradient boosting, 
* linear discriminant analysis,
* support vector machine.

```{r,results='hide',message=FALSE,warning=FALSE,cache=TRUE}
set.seed(125195)
model_RF <- randomForest(classe ~ ., data=trnSet_trn, importance = TRUE)
model_DT <- rpart(classe ~ ., data=trnSet_trn, method="class")
model_gbm <- train(classe ~ ., data=trnSet_trn, method="gbm")
model_lda <- train(classe ~ ., data=trnSet_trn, method="lda")
model_svm <- svm(classe ~ ., data=trnSet_trn)
```

# Accuracy of the models
To estimate out of sample error (models were built on `trnSet_trn`) we'll use `trnSet_tst`.

### Random forest model testing results

```{r}
prediction_RF <- predict(model_RF, trnSet_tst)
ConMx <- confusionMatrix(prediction_RF, trnSet_tst$classe)
print(ConMx)
head(apply(-varImp(model_RF),2,order))
plot(model_RF)
```

### Decision tree model testing results

```{r}
prediction_DT <- predict(model_DT, trnSet_tst, type="class")
ConMx2 <- confusionMatrix(prediction_DT, trnSet_tst$classe)
print(ConMx2)
order(-varImp(model_DT))
```
### Stochastic Gradient Boosting model testing results

```{r}
prediction_gbm <- predict(model_gbm, trnSet_tst)
ConMx3 <- confusionMatrix(prediction_gbm, trnSet_tst$classe)
print(ConMx3)
```
### Linear Discriminant Analysis model testing results

```{r}
prediction_lda <- predict(model_lda, trnSet_tst)
ConMx4 <- confusionMatrix(prediction_lda, trnSet_tst$classe)
print(ConMx4)
```
### Support vector machine model testing results

```{r}
prediction_svm <- predict(model_svm, trnSet_tst)
ConMx5 <- confusionMatrix(prediction_svm, trnSet_tst$classe)
print(ConMx5)
```

# Conclusions 
In out of sample error the RF model demonstrates the highest accuracy (`r round(ConMx$overall['Accuracy'],2)`). Also high accuracy was in gbm and svm models (`r round(ConMx3$overall['Accuracy'],2)` and `r round(ConMx5$overall['Accuracy'],2)`). Lowest accuracy was in lda (`r round(ConMx4$overall['Accuracy'],2)`) and rpart (`r round(ConMx2$overall['Accuracy'],2)`)


# Final test predictions

```{r}
prediction <- rbind(predict(model_RF, tstSet),predict(model_DT, tstSet,type="class"),
                    predict(model_gbm, tstSet),predict(model_lda, tstSet),
                    predict(model_svm, tstSet))
Pred_Fin <- as.data.frame(x=matrix(c("A","B","C","D","E")[prediction], ncol = 20, nrow = 5),
                          row.names = c("RF","DT","gbm","lda","svm"))
print(Pred_Fin)
```
For final quiz we'll use the results of the random forest model which demonstrate the best accuracy (`r round(ConMx$overall['Accuracy'],2)`).